Virtualization of a Real-Time Operating System for Robot Control with a Focus on Real-Time Compliance

### 2024-09-02 Masterarbeit_Robotik_Pamuk.tex
Fertiggggg

\bigskip \noindent As a next step, the robotic application in a practical scenario also demonstrates and validates the improvement of the tuned virtualization. The difference between the command time and the time the signal reaches the PWM module was measured 1,000 times for the untuned, tuned, and bare metal versions of Salamander 4. The most important metric here is also the worst latency of the signal reaching the PWM. It dropped from 129~ms to 3.988~ms in the developed application. Comparing this to the worst latency value of the bare metal version, which is 1.49~ms, it is apparent that the tuned virtualization through real-time performance configurations came very close to the determinism and reliability of the hardware.

@@

\bigskip \noindent On top of that, a robotic application further confirmed the improvements. The difference between the command time and the time the signal reaches the PWM module was measured 1,000 times for the untuned, tuned, and bare metal versions of Salamander 4. The worst latency of the signal reaching the PWM dropped from an initial value of 129~ms to 3.988~ms. Comparing this to the worst latency value of the bare metal version, which was 1.49~ms, it is apparent that the tuned virtualization through real-time performance configurations came very close to the determinism and reliability of the hardware.

>>>>

\bigskip \noindent As a next step, the robotic application in a practical scenario also demonstrates and validates the improvement of the tuned virtualization. The difference between the command time and the time the signal reaches the PWM module was measured 1,000 times for the bare metal, untuned, and tuned versions of Salamander 4. The most important metric here is also the worst latency of the signal reaching the PWM. It dropped from 129~ms to 3.988~ms in the developed application. Comparing this to the worst latency value of the bare metal version, which is 1.49~ms, it is apparent that the tuned virtualization through real-time performance configurations came very close to the determinism and reliability of the hardware.

@@

\bigskip \noindent On top of that, a robotic application further confirmed the improvements. The difference between the command time and the time the signal reaches the PWM module was measured 1,000 times for the bare metal, untuned, and tuned versions of Salamander 4. The worst latency of the signal reaching the PWM dropped from an initial value of 129~ms to 3.988~ms. Comparing this to the worst latency value of the bare metal version, which was 1.49~ms, it is apparent that the tuned virtualization through real-time performance configurations came very close to the determinism and reliability of the hardware.

### 2024-09-02 Paper.tex
Fertiggggg

  \section{Methodology}
  In order to address this gap and achieve determinism, an extensive real-time performance tuning process was carried out. These tunings were added sequentially and tested together. This involves configurations spanning the BIOS, kernel, host operating system (OS), and QEMU/KVM virtualization layer, as demonstrated in~\cite{RealTimePerformanceTuning2022}. The guest OS is already based on Xenomai and therefore, no additional modifications were necessary. Ubuntu 22.04.4 LTS served as the host operating system, while Salamander 4, built with Yocto and based on Linux 5.15.94, was the guest OS. Xenomai provided real-time capabilities for the guest and its \texttt{latency} tool was used to measure the scheduling latency of the real-time OS. QEMU with KVM was employed for virtualization. Trace-cmd and Kernelshark were used for kernel tracing and visualization of trace data. The testing robot was connected to the virtual CPU via a VARAN bus interface.

  \section{Practical implementation}
  The initial latency values were first measured with the \texttt{latency} program of the Xenomai tool suite. The tests were conducted for 10 minutes with a sampling period of 100~µs, using a periodic user-mode task, and were assigned a priority of 99. If any sample's latency value was greater than 100~µs, this was considered an overrun. There was a significant initial gap in latency between the bare metal and virtualized Salamander 4. To address this gap, an extensive tuning process was carried out to achieve real-time performance and determinism. These modifications were added sequentially and tested together. Previous tunings were not reverted when moving to the next tunings. First, the BIOS was configured, followed by the kernel. The BIOS settings were not reverted when moving to the kernel configurations. Next, the host was configured, but the BIOS and kernel settings remained unchanged. The guest configurations were not changed since Xenomai already provides real-time capabilities. Finally, QEMU/KVM settings were applied. On top of that, a robotic application further confirmed the improvements. The difference between the command time and the time the signal reaches the Pulse Width Modulation (PWM) Module  was measured 1,000 times to get reliable results for the bare metal, untuned, and tuned versions of Salamander 4.

@@

  Figure~\ref{fig:max_latency_combined_results} displays the change in latency after each real-time performance tuning.

@@

  \begin{table}[H]
    \centering
    \footnotesize
    \caption{Comparison of Latency Statistics in Salamander 4 after Configurations}
    \label{tab:latency_tables_combined}
    \begin{tabular}{|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{\textbf{Version}} & \multicolumn{3}{c|}{\textbf{Latency (µs)}} & \multirow{2}{*}{\textbf{Overruns}} \\ \cline{2-4}
      & \textbf{Min} & \textbf{Avg} & \textbf{Max} & \\ \hline
      Bare Metal & 0.613 & 1.380 & 10.709 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}Untuned\\ Virtualization\end{tabular} & 2.536 & 8.940 & 707.622 & 43 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After BIOS\\ Configurations\end{tabular} & 0.969 & 3.948 & 457.545 & 94 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After Kernel\\ Configurations\end{tabular} & 2.545 & 4.811 & 21.694 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After Host\\ Configurations\end{tabular} & 2.591 & 4.834 & 18.441 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After QEMU\\ Configurations\end{tabular} & 2.614 & 4.779 & 17.134 & 0 \\ \hline
    \end{tabular}
\end{table}

@@

  \section{Discussion}
  The results show that the real-time performance tunings applied to the virtualized Salamander 4 OS reduced latency and improved determinism. The maximum latency decreased from 707.622~µs to 17.134~µs after tuning BIOS, kernel, host, and QEMU/KVM configurations. Especially the BIOS and kernel configurations played a crucial role in reducing the maximum latency, with a major fall to 21.694~µs. This also eliminated overruns going forward. The host configurations, including CPU and interrupt affinity, and real-time prioritization of QEMU, further reduced latency down to 18.441~µs. The guest OS already had real-time capabilities through Xenomai, hence the focus was primarily on optimizing the host and virtualization layer to achieve the desired real-time performance. Finally, QEMU/KVM configurations, such as tuning the LAPIC timer advance and using hugepages, brought the latency down to the final worst latency value of 17.134~µs. This value is very close to the bare metal value of 10.709~µs. The goal was set to achieving latency values below 50~µs in the duration of the measurement. This goal was achieved.
  A robotic application in a practical scenario validated the improvement of the tuned virtualization compared to the unmodified version. The difference between command time and signal reaching the PWM module was measured 1,000 times for the bare metal, untuned, and tuned versions of Salamander 4. The worst latency dropped from 129~ms to 3.988~ms in the developed application. Compared to the bare metal version's worst latency of 1.49~ms, the tuned virtualization came very close to the determinism and reliability of the bare metal.

  \section{Summary and Outlook}
  This master's thesis aimed to virtualize a real-time operating system for robot control, focusing on real-time determinism. The goal was to reduce latency in virtualized Salamander 4 to match its bare metal version. Initial tests showed a significant latency gap between bare metal and virtualized versions. Extensive tuning was performed sequentially, including BIOS, kernel, host, and QEMU/KVM configurations. This process reduced worst latency from 707.62~µs to 17.134~µs, close to the bare metal value of 10.709~µs, achieving the goal of sub-50~µs latency.
  A robotic application confirmed these improvements. Measuring the time between command issuance and command arrival at the PWM module 1,000 times showed worst latency dropping from 129~ms to 3.988~ms, approaching the bare metal version's 1.49~ms.  
  Future work could include additional configurations and optimizations of the virtualization layer, investigating other hypervisors and virtualization technologies, and extensive testing under various workloads and stressed situations.

>>>>

  \section{Methodology}
  In order to address this gap and achieve determinism, an extensive real-time performance tuning process was carried out. These tunings were added sequentially and tested together. This involves configurations spanning the BIOS, kernel, host operating system (OS), and QEMU/KVM virtualization layer, as demonstrated in~\cite{RealTimePerformanceTuning2022}. Xenomai provided real-time capabilities for the guest OS and therefore, no additional modifications were necessary. Ubuntu 22.04.4 LTS served as the host operating system, while Salamander 4, built with Yocto and based on Linux 5.15.94, was the guest OS. The \texttt{latency} tool of the Xenomai tool suite was used to measure the scheduling latency of the real-time OS. QEMU with KVM was employed for virtualization. Trace-cmd and Kernelshark were used for kernel tracing and visualization of trace data. The testing robot was connected to the virtual CPU via a VARAN bus interface.

  \section{Practical implementation}
  The initial latency values were first measured with the \texttt{latency} program of the Xenomai tool suite. The tests were conducted for 10 minutes with a sampling period of 100~µs, using a periodic user-mode task, and were assigned a priority of 99. If any sample's latency value was greater than 100~µs, this was considered an overrun. There was a significant initial gap in latency between the bare metal and virtualized Salamander 4. To address this gap, an extensive tuning process was carried out to achieve real-time performance and determinism. These modifications were added sequentially and tested together. Previous tunings were not reverted when moving to the next tunings. First, the BIOS was configured, followed by the kernel. The BIOS settings were not reverted when moving to the kernel configurations. Next, the host was configured, but the BIOS and kernel settings remained unchanged. The guest configurations were not changed since Xenomai already provides real-time capabilities. Finally, QEMU/KVM settings were applied. A robotic application further confirmed the improvements. The difference between the command time and the time the signal reaches the Pulse Width Modulation (PWM) Module  was measured 1,000 times to get reliable results for the bare metal, untuned, and tuned versions of Salamander 4.

@@

  Figure~\ref{fig:max_latency_combined_results} displays the change in latency after each real-time performance tuning of Salamander 4.

@@

  \begin{table}[H]
    \centering
    \footnotesize
    \caption{Comparison of Latency Statistics in Salamander 4 after Configurations}
    \label{tab:latency_tables_combined}
    \begin{tabular}{|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{\textbf{Version}} & \multicolumn{3}{c|}{\textbf{Latency (µs)}} & \multirow{2}{*}{\textbf{Overruns}} \\ \cline{2-4}
      & \textbf{Min} & \textbf{Avg} & \textbf{Max} & \\ \hline
      Bare Metal & 0.613 & 1.380 & 10.709 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}Untuned\\ Virtualization\end{tabular} & 2.536 & 8.940 & 707.622 & 43 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After BIOS\\ Configurations\end{tabular} & 0.969 & 3.948 & 457.545 & 22 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After Kernel\\ Configurations\end{tabular} & 2.545 & 4.811 & 21.694 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After Host\\ Configurations\end{tabular} & 2.591 & 4.834 & 18.441 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After QEMU\\ Configurations\end{tabular} & 2.614 & 4.779 & 17.134 & 0 \\ \hline
    \end{tabular}
\end{table}

@@

  \section{Discussion}
  The results show that the real-time performance tunings applied to the virtualized Salamander 4 OS reduced latency and improved determinism. The maximum latency decreased from 707.622~µs to 17.134~µs after tuning BIOS, kernel, host, and QEMU/KVM configurations. Especially, applying the PREEMPT-RT patch, along with optimizing both the BIOS and kernel configurations, played a crucial role in reducing the maximum latency to 21.694~µs. This also eliminated overruns going forward. The host configurations, including CPU and interrupt affinity, and real-time prioritization of QEMU, further reduced latency down to 18.441~µs. The guest OS already had real-time capabilities through Xenomai, hence the focus was primarily on optimizing the host and virtualization layer to achieve the desired real-time performance. Finally, QEMU/KVM configurations, such as tuning the LAPIC timer advance and using hugepages, brought the latency down to the final worst latency value of 17.134~µs. This value is very close to the bare metal value of 10.709~µs. The goal was set to achieving latency values below 50~µs in the duration of the measurement. This goal was achieved.
  A robotic application in a practical scenario validated the improvement of the tuned virtualization compared to the unmodified version. The difference between the command time and the time the signal reaches the PWM module was measured 1,000 times for the bare metal, untuned, and tuned versions of Salamander 4. The worst latency in the virtualization dropped from 129~ms to 3.988~ms in the developed application. Compared to the bare metal version's worst latency of 1.49~ms, the tuned virtualization came very close to the determinism and reliability of the bare metal.

  \section{Summary and Outlook}
  This master's thesis aimed to virtualize an RTOS for robot control, focusing on real-time determinism. The goal was to reduce latency in the virtualized Salamander 4 to match its bare metal version. Initial tests showed a significant latency gap between bare metal and virtualized versions. Extensive tuning was performed sequentially, including BIOS, kernel, host, and QEMU/KVM configurations. This process reduced the worst latency from 707.62~µs to 17.134~µs, close to the bare metal value of 10.709~µs, achieving the goal of sub-50~µs latency.
  A robotic application confirmed these improvements. Measuring the time between command issuance and command arrival at the PWM module 1,000 times showed worst latency dropping from 129~ms to 3.988~ms, approaching the bare metal version's 1.49~ms.  
  Future work could include additional configurations and optimizations of the virtualization layer, investigating other hypervisors and virtualization technologies, and extensive testing under various workloads and stressed situations.
